All Your Graphs Belong to Us

### 2012-07-18 doc/asplos13/implementation.tex
Rearranged implementation and made a pass on it.

@@

Grappa tasks are 32-byte entities: a 64-bit function pointer plus
three 64-bit arguments. We use three arguments because tasks are
usually generated as part of a parallel loop decomposition, and thus
each task needs three kinds of data:
\begin{description}
\item[Function pointer] This 64-bit value indicates what routine
  should run. We depend on the fact that all processes in the system
  load the same binary so that function pointers are consistent across
  processes.
\item[Private argument] This 64-bit value is used for task-specific
  data, and is often a loop index.
\item[Shared argument] This 64-bit value is used for data shared
  between all tasks that are part of a loop, and is usually a global
  pointer to a struct. Each task uses its private argument to 
\item[Synchronization argument] This 64-bit value is used to determine
  when all tasks that are part of a loop have finished. It is usually
  a global pointer to a synchronization object allocated at the core
  that spawned the task.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitray 64-bit values in the runtime, and can be used
to store whatever the user wants.

Tasks are not allocated any execution resources until the scheduler
decides to run them; when this occurs, tasks are matched with {\em
  workers}. Each worker is simply a collection of status bits and a
stack, allocated at a particular core. Workers

@@

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we can declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch. In
particular, only one of our applications does floating point
operations; all the other applications can safely ignore the x87 and
SSE2 registers as part of context switching.

@@

@@

The scheduler uses its core's timestamp counter to track time. The
timestamp counter is read once per scheduling decision. Timestamp
values are not shared between cores. We use timestamp values only to
ensure period execution of communication maintenance tasks; being
wrong will only change the rate of these events with a potential
performance cost, and no correctness violation. Nevertheless, we
assume Grappa will run on machines with modern timestamp counter
implementations that are monotonic and frequency-compensated.

@@

Most study of work stealing assumes that a processor has one worker
thread and so it only steals when utilization would go to zero. 
%that is,when there are no more local tasks available. 
Grappa, on the other hand, relies on having many worker threads per core. The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized.
So, there is a choice: should the core use some of this idle time to
do perform work stealing or should it just wait that local tasks will
create more work. We choose to have a core inititiate a steal when all
of the following conditions hold: no workers are ready to run, 
%\checkme{there are idle worker threads --not required to mention since
%technically would happen if have right number}
the unstarted task queues are empty, and there are no outstanding steal requests. 
Having one outstanding steal request allowed throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as voting tree\cite{scalableWorkStealingOrCilk98}
or lifelines \cite{lifelines}. Termination detection is not
built into the Grappa task scheduler, rather, it is considered a programming error for
the program to exit without syncing all tasks.

@@

In order to mitigate the low message injection rate limits of commodity
networks, Grappa's communication has two layers.

@@

Underneath the aggregation layer, Grappa uses the GASNet communication
library \TODO{cite GASNet} to actually move data. All interprocess
communication, whether on or off a cluster node, is handled by the
GASNet library. GASNet is able to take advantage of many communication
mechanism, including Ethernet and Infiniband between nodes, as well as
shared memory within a node.

Some networks provide access to a remote machine's memory
directly. This would seem to be a good fit for a programming model
focused on global shared memory, but in fact we do not use it. RDMA
operations are subject to the same message rate limitations as all
other messages on these cards, and thus using raw RDMA operations for
our small messages would make inefficient use of bandwidth. Instead,
we implement remote memory operations with active messages as
described in the next section. This means that Grappa is not limited
to RDMA-capable networks.

All global addresses in Grappa are 64-bit values. Each byte of global
memory is associated with a {\em home core}; this core is responsible
for allocating and modifying that memory. Since today's commodity
processors support a 48-bit virtual address space \TODO{cite?}, Grappa
uses the spare bits to distinguish which address space a global
address belongs to, and to represent the home core of a global
address.

The second global address space is the {\em linear} address space, so
named because nearby addresses can be treated as adjacent even if they
are stored on different nodes. Addresses in the linear address space
are partitioned across the cluster in a block-cyclic fashion. Choosing
the block size involves trading off sequential bandwidth agains
aggregate random access bandwidth. Smaller block sizes help spread
data across all the memory controllers in the cluster, but larger
block sizes allow the locality-optimized memory controllers to provide
increased sequential bandwidth. The block size is typically set to 64
bytes, or the size of a single hardware cache line, in order to
exploit spatial locality when available.

Delegate operations are used for syncronization. Their modifications
are always executed at the home core of their address. Each task may
have only one outstanding delegate operation. We limit delegate
operations to operate objects in the 2D address space or objects that
fit in a single block of the linear address space. No context switches
are allowed while the data it being modified. Given these
restrictions, we can ensure that delegate operations from multiple
requestors are always serialized through a single core in the system,
providing atomic semantics without using atomic operations.

All current cache operations are incoherent; two tasks may
concurrently modify the same block with their writes interfereing
unless protected by synchronization through delegate operations. While
this has been sufficient for our benchmarks, it is possible to embed
synchronization in the cache active messages and reduce the number of
network round trips necessary.

>>>>

\paragraph{2D global addresses:} All global addresses in Grappa are 64-bit values. Each byte of global
memory is associated with a {\em home core}; this core is responsible
for allocating and modifying that memory. Since today's commodity
processors support a 48-bit virtual address space \TODO{cite?}, Grappa
uses the spare bits to distinguish which address space a global
address belongs to, and to represent the home core of a global
address.

\paragraph{Linear global addresses:} The second global address space is the {\em linear} address space, so
named because nearby addresses can be treated as adjacent even if they
are stored on different nodes. Addresses in the linear address space
are partitioned across the cluster in a block-cyclic fashion. Choosing
the block size involves trading off sequential bandwidth against
aggregate random access bandwidth. Smaller block sizes help spread
data across all the memory controllers in the cluster, but larger
block sizes allow the locality-optimized memory controllers to provide
increased sequential bandwidth. The block size, which is configurable, is typically set to 64
bytes, or the size of a single hardware cache line, in order to
exploit spatial locality when available.

\TODO{Add a figure that depicts both forms of global addresses.}

\paragraph{Delegates:} Delegate operations are used for syncronization. Their modifications
are always executed at the home core of their address. Each task may
have only one outstanding delegate operation. We limit delegate
operations to operate on objects in the 2D address space or objects that
fit in a single block of the linear address space. No context switches
are allowed while the data it being modified. Given these
restrictions, we can ensure that delegate operations for the same address from multiple
requestors are always serialized through a single core in the system,
providing atomic semantics without using atomic operations.

\comment{If anything, this is just a footnote. -M

All current cache operations are incoherent; two tasks may
concurrently modify the same block with their writes interfereing
unless protected by synchronization through delegate operations. While
this has been sufficient for our benchmarks, it is possible to embed
synchronization in the cache active messages and reduce the number of
network round trips necessary.}

@@

Grappa tasks are 32-byte entities: a 64-bit function pointer plus
three 64-bit arguments. We use three arguments because tasks are
usually generated as part of a parallel loop decomposition, and thus
each task needs three kinds of data:
\begin{description}
\item[Function pointer] This 64-bit value indicates what routine
  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
\item[Private argument] This 64-bit value is typically used for task-specific
  data (e.g., a loop index).
\item[Shared argument] This 64-bit value is typically used for data shared
  between all tasks that are part of a loop.
\item[Synchronization argument] This 64-bit value is typically used to determine
  when all tasks that are part of a loop have finished. It is usually
  a global pointer to a synchronization object allocated at the core
  that spawned the task.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitray 64-bit values in the runtime, and can be used
for any purpose.

Tasks are not allocated any execution resources until the scheduler
decides to run them; when this occurs, tasks are matched with {\em
  workers}. Each worker is simply a collection of status bits and a
stack, allocated at a particular core.

@@

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.
\comment{I highly doubt this is important in real apps beyond our kernels.  -M

In
particular, only one of our applications does floating point
operations; all the other applications can safely ignore the x87 and
SSE2 registers as part of context switching.}

@@

\TODO{What is a worker?  We don't realyl describe this.  -M}

@@

The scheduler uses its core's timestamp counter to track time. The
timestamp counter is read once per scheduling decision. Timestamp
values are not shared between cores. We use timestamp values only to
ensure periodic execution of communication maintenance tasks; being
wrong will only change the rate of these events with a potential
performance cost, but does not cause a correctness violation. Nevertheless, we
assume Grappa will run on machines with modern timestamp counter
implementations that are monotonic and frequency-compensated.

@@

Most study of work stealing assumes that a processor has one worker
thread and so it only steals when utilization would go to zero. 
Grappa, on the other hand, relies on having many worker threads per core. The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized.
So, there is a choice: should the core use some of this idle time to
perform work stealing or should it just wait as it is possible that local tasks will
create more work. We choose to have a core inititiate a steal when all
of the following conditions hold: no workers are ready to run, 
the unstarted task queues are empty, and there are no outstanding steal requests. 
Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting tree\cite{scalableWorkStealingOrCilk98}
or lifelines \cite{lifelines}. Termination detection is not
built into the Grappa task scheduler, rather, it is considered a programming error for
the program to exit without syncing all tasks.

@@

In order to mitigate the low message injection rate limits of commodity
networks, Grappa's communication stack has two layers.

@@

Underneath the aggregation layer, Grappa uses the GASNet communication
library \TODO{cite GASNet} to actually move data. All interprocess
communication, whether on or off a cluster node, is handled by the
GASNet library. GASNet is able to take advantage of many communication
mechanisms, including Ethernet and Infiniband between nodes, as well as
shared memory within a node.

Some networks provide access to a remote machine's memory
directly. This would seem to be a good fit for a programming model
focused on global shared memory, but in fact we do not use it. RDMA
operations are subject to the same message rate limitations as all
other messages on these cards, and thus using raw RDMA operations for
our small messages would make inefficient use of bandwidth. Instead,
we implement remote memory operations with active messages. A byproduct of this design decision is that Grappa is not limited
to RDMA-capable networks.
